Two subtle problems with over-representation analysis

### 2024-06-14 manuscript/manuscript.Rmd
mswork

To test the effect of these two problems on RNA-seq based enrichment analysis, data from seven
transcriptomic studies were downloaded from DEE2 using the getDEE2 R/Bioconductor package
[@Ziemann2019-ux].
Raw data for these seven studies are available from NCBI Sequence Read Archive under the
accession numbers SRP128998, SRP038101, SRP037718, SRP096177, SRP097759, SRP253951 and
SRP068733 [@Felisbino2021-hc; @Lund2014-ds; @Rafehi2014-rv; @Keating2014-ab;
@Blanco-Melo2020-hh; @Rafehi2017-qn].
For each dataset, kallisto (v0.43.1) [@Bray2016-lu] transcript counts were aggregated to the
gene level.
Genes with fewer than 10 reads per sample on average were removed from downstream analysis.
Remaining genes passing this selection were included in the background gene set.
Differential expression analysis was conducted with DESeq2 v1.42.0 [@Love2014-wg], with
genes identified by their Ensembl identifiers.
Gene symbols were then fetched using biomaRt v2.58.0, based on Ensembl version 112 [@Durinck2009-dr].

@@

Human gene sets were obtained from MSigDB v2023.2 [@Liberzon2015-cs], and included
KEGG Medicus [@Kanehisa2023-ni],
Reactome [@Milacic2024-iv],
Wikipathways [@Agrawal2024-lr],
MicroRNA targets from miRdb [@Chen2020-wo],
transcription factor targets from GTRD [@Kolmykov2021-mn],
Gene Ontology (GO) [@The_Gene_Ontology_Consortium2023-xp],
Human Phenotype Ontology (HPO) [@Kohler2021-ha],
Cell markers and
Hallmark pathways [@Liberzon2015-cs].
To demonstrate the background problem, differentially expressed genes with
FDR<0.05 were subset, with separate lists for up and downregulated genes.
If fewer than 200 genes met this criteria, then the 200 genes with the smallest p-values
were selected, as suggested by a previous report [@Tarca2013-xh].
These gene lists were subjected to ORA using clusterProfiler's `enricher` function
v4.10.0, using a minimum set size of 5 and a no maximum set size.

@@

RNA-seq counts for SRA run accession ERR2539161 were obtained from DEE2 [@Mo2018-zb].
This dataset was selected due to its high number of assigned reads (367M).
A library of 200 random gene sets were generated, each containing 30 genes.
To generate differential expression profiles, 5% of gene sets were selected to be
differentially expressed, with equal numbers of up- and down-regulated pathways.
Pseudosamples were generated by first downsampling the counts using the `thincounts`
function of the `edgeR` package v4.2.0 [@Robinson2010-eb] to 20M reads, followed by the
addition of some extra variation drawn from a normal distribution with mean of 1 and
standard deviation varying between 0.1 and 0.6, this value selected randomly for each gene.
Three "control" and five "case" pseudosamples were generated.
For the genes selected to be differentially expressed, expression values were multiplied by a
log fold change in the case samples of 0.5 for upregulated genes and -0.5 for downregulated
genes.
Genes that were selected to be both up and downregulated were left unchanged.
With these expression profiles, DESeq2 was used for differential analysis, and the results
were subjected to clusterProfiler which suffers from both problems,
clusterProfiler with the background problem fix, clusterProfiler with the FDR problem fix,
clusterProfiler with the fixes to background and FDR problems, an ORA function called `fora`
from `fgsea` bioconductor package v1.30.0 that does not suffer these problems, and `fgsea` a
fast implementation of the preranked gene set enrichment analysis (GSEA) algorithm
[@Korotkevich2016-gd].
For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
selected for the foreground gene set.
If fewer than 200 genes were detected as significant, then 200 genes with the smallest
p-values were used for ORA, similar to a previous report [@Tarca2013-xh].
For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.
After enrichment analysis with these approaches, the significant up and down regulated
sets were selected and compared to the ground truth in order to calculate precision, recall
and F1 score.
At each value of added variance, 1000 replications were conducted.
No set seed was used, but results were similar when run on different computer systems.
All analyses were undertaken with R v4.4.0.

@@

A workaround to eliminate the background problem was developed and we used it to compare the
results of ORA of seven datasets with the original and corrected methods for nine different
gene set libraries.
The mean number of significant results was calculated for each of the gene set libraries
before and after correcting the problem (**Figure 1A**).
Each of the gene set libraries showed an uplift when correcting the background problem.
The smallest increase was observed in Cellmarkers gene library (11.7%), while the largest
increase was observed for HPO (229%).
The similarity between results, as quantified with the Jaccard similarity was highest for
Cellmarkers and GTRD at 0.88 and 0.76 respectively, and the lowest similarity was recorded for
HPO and KEGG with 0.20 and 0.48 respectively.
Observed Jaccard scores correlated with the number of genes annotated to one or more functional
categories in the gene set library (**Figure 1B**).
This result suggests that ORA with smaller gene set libraries like HPO and KEGG are impacted
more severely.

@@

Simulated differential expression profiles with *a priori* changes to predetermined pathway
genes were analysed in parallel by clusterProfiler default and with mitigations for
problems described above.
In addition, these data underwent ORA with fora which does not suffer from these two
problems.
For comparison fgsea was used to compare the performance of these two ORA methods against a
prototypical FCS method.
These methods were tested with different levels of introduced variation and the mean
precision, recall and F1 scores were recorded (**Figure 3A**).
As anticipated, mitigating the background problem improved recall (by 1.6%), however it caused
a similar-sized decrease in precision.
Addressing the FDR problem improved precison by 9.6%, but the recall dropped by 8.5%.
Addressing both problems improved precision by 8.2% but decreased recall by 5.6%.
Under these conditions, the FDR problem affected results more severely than the background
problem.
Results obtained for clusterProfiler after implementing the two mitigations were identical to
fora to three significant figures.
fgsea recorded precision of 0.937, similar to the ORA methods, but recall was far superior,
with fgsea scoring 22.4% higher than default clusterProfiler and 28% higher than fora.
As a result, fgsea's overall accuracy as defined by F1 index was 21.2% higher than fora and
18.4% higher than default clusterProfiler.

When looking at the accuracy profile at different levels of added noise (**Figure 3B**), we
see that ORA techniques have superior precision when the added noise is low (<0.2), but the
precision of ORA techniques drops at higher levels of added noise.
The rate of drop in precision was more severe for approaches that suffer from the FDR problem.
The precision of fgsea was relatively stable despite high levels of added noise, ranging from
0.95 to 0.91.
The recall of fgsea was superior to all ORA approaches across all levels of added noise.

@@

The background problem came to our attention when a doctoral student in our department
was puzzled by results obtained from clusterProfiler when using a custom set of pathways.
This exact issue was the subject of several posts to online bioinformatics forums dating to
~2020 where confused users were wondering why the number of background genes changed so much.
The results shown here show that the background problem effects smaller gene set libraries
like KEGG and HPC more severely as compared to Cellmarkers and GO libraries which describe
a larger proportion of all genes.
ClusterProfiler maintainers recently provided an option to prevent the loss of unannotated
genes from the background `options(enrichment_force_universe = TRUE)`, however this feature
isn't yet described in the official documentation.

>>>>

To test the effect of these two problems on RNA-seq based enrichment analysis, data from seven
transcriptomic studies were downloaded from DEE2 using the getDEE2 R/Bioconductor package
[@Ziemann2019-ux].
Raw data for these seven studies are available from NCBI Sequence Read Archive under the
accession numbers SRP128998, SRP038101, SRP037718, SRP096177, SRP097759, SRP253951 and
SRP068733 [@Felisbino2021-hc; @Lund2014-ds; @Rafehi2014-rv; @Keating2014-ab;
@Blanco-Melo2020-hh; @Rafehi2017-qn].
For each dataset, kallisto (v0.43.1) [@Bray2016-lu] transcript counts were aggregated to the
gene level.
Genes with fewer than 10 reads per sample on average were removed from downstream analysis.
Remaining genes passing this selection were included in the background gene set.
Differential expression analysis was conducted with DESeq2 v1.44.0 [@Love2014-wg], with
genes identified by their Ensembl identifiers.
Gene symbols were then fetched using biomaRt v2.60.0, based on Ensembl version 112 [@Durinck2009-dr].

@@

Human gene sets were obtained from MSigDB v2023.2 [@Liberzon2015-cs], and included
KEGG Medicus [@Kanehisa2023-ni],
Reactome [@Milacic2024-iv],
Wikipathways [@Agrawal2024-lr],
MicroRNA targets from miRdb [@Chen2020-wo],
transcription factor targets from GTRD [@Kolmykov2021-mn],
Gene Ontology (GO) [@The_Gene_Ontology_Consortium2023-xp],
Human Phenotype Ontology (HPO) [@Kohler2021-ha],
Cell markers and
Hallmark pathways [@Liberzon2015-cs].
To demonstrate the background problem, differentially expressed genes with
FDR<0.05 were subset, with separate lists for up and downregulated genes.
If fewer than 200 genes met this criteria, then the 200 genes with the smallest p-values
were selected, as suggested by a previous report [@Tarca2013-xh].
These gene lists were subjected to ORA using clusterProfiler's `enricher` function
v4.12.0, using a minimum set size of 5 and a no maximum set size.

@@

RNA-seq counts for SRA run accession ERR2539161 were obtained from DEE2 [@Mo2018-zb].
This dataset was selected due to its high number of assigned reads (367M).
A library of 200 random gene sets were generated, each containing 30 genes.
To generate differential expression profiles, 5% of gene sets were selected to be
differentially expressed, with equal numbers of up- and down-regulated pathways.
Pseudosamples were generated by first downsampling the counts using the `thincounts`
function of the `edgeR` package v4.2.0 [@Robinson2010-eb] to 20M reads, followed by the
addition of some extra variation drawn from a normal distribution with mean of 1 and
standard deviation varying between 0.1 and 0.6, this value selected randomly for each gene.
Three "control" and five "case" pseudosamples were generated.
For the genes selected to be differentially expressed, expression values were multiplied by a
log fold change in the case samples of 0.5 for upregulated genes and -0.5 for downregulated
genes.
Genes that were selected to be both up and downregulated were left unchanged.
With these expression profiles, DESeq2 was used for differential analysis, and the results
were subjected to clusterProfiler which suffers from both problems,
clusterProfiler with the background problem fix, clusterProfiler with the FDR problem fix,
clusterProfiler with the fixes to background and FDR problems, an ORA function called `fora`
from `fgsea` bioconductor package v1.30.0 that does not suffer these problems, and `fgsea` a
fast implementation of the preranked gene set enrichment analysis (GSEA) algorithm
[@Korotkevich2016-gd].
For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
selected for the foreground gene set.
If fewer than 200 genes were detected as significant, then 200 genes with the smallest
p-values were used for ORA, similar to a previous report [@Tarca2013-xh].
For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.
After enrichment analysis with these approaches, the significant up and down regulated
sets were selected and compared to the ground truth in order to calculate precision, recall
and F1 score.
At each value of added variance, 1000 replications were conducted.
No set seed was used, but results were similar when run on different computer systems.
All analyses were undertaken with R v4.4.0 in a Docker container.

@@

A workaround to eliminate the background problem was developed and we used it to compare the
results of ORA of seven datasets with the original and corrected methods for nine different
gene set libraries.
The mean number of significant results was calculated for each of the gene set libraries
before and after correcting the problem (**Figure 1A**).
Each of the gene set libraries showed an uplift when correcting the background problem.
The smallest increase was observed in Cellmarkers gene library (11.8%), while the largest
increase was observed for HPO (224%).
The similarity between results, as quantified with the Jaccard similarity was highest for
Cellmarkers and GTRD at 0.76 and 0.77 respectively, and the lowest similarity was recorded for
HPO and KEGG with 0.20 and 0.48 respectively.
Observed Jaccard scores correlated with the number of genes annotated to one or more functional
categories in the gene set library (**Figure 1B**).
This result suggests that ORA with smaller gene set libraries like HPO and KEGG are impacted
more severely.

@@

Simulated differential expression profiles with *a priori* changes to predetermined pathway
genes were analysed in parallel by clusterProfiler default and with mitigations for
problems described above.
In addition, these data underwent ORA with fora which does not suffer from these two
problems.
For comparison fgsea was used to compare the performance of these two ORA methods against a
prototypical FCS method.
These methods were tested with different levels of introduced variation and the mean
precision, recall and F1 scores were recorded (**Figure 3A**).
Mitigating the background problem improved recall (by 1.8%), however it caused
a similar-sized decrease in precision (1.6% lower).
Addressing the FDR problem improved precison by 9.9%, but the recall dropped by 7.7%.
Addressing both problems improved precision by 8.2% but decreased recall by 5.4%.
Under these conditions, the FDR problem affected results more severely than the background
problem.
Results obtained for clusterProfiler after implementing the two mitigations were identical to
fora to three significant figures.
fgsea recorded precision of 0.940, similar to the ORA methods, but recall was far superior,
with fgsea scoring 22.9% higher than default clusterProfiler and 28.3% higher than fora.
As a result, fgsea's overall accuracy as defined by F1 index was 21.0% higher than fora and
18.6% higher than default clusterProfiler.

When looking at the accuracy profile at different levels of added noise (**Figure 3B**),
ORA techniques show high precision when the added noise is low (<0.2), but
precision drops at higher levels of added noise.
The rate of drop in precision was more severe for approaches that suffer from the FDR problem.
The precision of fgsea was relatively stable despite high levels of added noise, ranging from
0.95 to 0.92.
The recall of fgsea was superior to all ORA approaches across all levels of added noise.

@@

The background problem came to our attention when a doctoral student in our department
was puzzled by results obtained from clusterProfiler when using a custom set of pathways.
This exact issue was the subject of several posts to online bioinformatics forums dating to
~2020 where confused users were wondering why the number of background genes changed so much.
The results shown here show that the background problem effects smaller gene set libraries
like KEGG and HPO more severely as compared to Cellmarkers and GO libraries which describe
a larger proportion of all genes.
ClusterProfiler maintainers recently provided an option to prevent the loss of unannotated
genes from the background `options(enrichment_force_universe = TRUE)`, however this feature
isn't yet described in the official documentation.
